---
title: "Minnesota Covid 19 data"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

# Introduction

What is the trend in Covid-19 cases in Minnesota?

# `r format(Sys.time(), '%B %e, %Y')`

```{r setup, include=FALSE}
library(tidyverse)
library(kableExtra)

knitr::opts_chunk$set(
  # Don't put code in the doc
  echo = FALSE,
  # Don't put messages from code in the doc
  message = FALSE,
  # Don't put code warnings in the doc
  warning = FALSE,
  # stop on error instead of continuing
  error = FALSE,
  # Don't let figures float
  # Needs "float" package, see https://stackoverflow.com/a/36234023/34935
  fig.pos = "H",
  # Default height 3.75 inches: two per page plus text
  fig.height = 3.75
)

theme_set(theme_bw() +
            theme(plot.title=element_text(size=25)))
```

```{r graphs, fig.show = "hold", out.width = "50%"}
########## hospitalized

hospdata <- read_tsv('data/hosptable.tsv')
names(hospdata) <- c('date', 'icu', 'non_icu',
                    'total_hosp', 'total_icu_hosp')
hospdata <- hospdata %>% mutate(
  date = as.Date(paste0("2020/", date), format="%Y/%m/%d")
)

# last 8 weeks
first_day <- max(hospdata$date) - 28*2

hospdata_long <- hospdata %>%
  pivot_longer(c('icu', 'non_icu'))
ggplot(hospdata_long %>% filter(date >= first_day),
       aes(date, value, group=name, color=name)) +
  geom_point() + geom_line() +
  geom_smooth() +
  labs(title="Hospitalized",
       subtitle="as of the given date",
       y="hospitalized as of given date")

########## deaths

deathdata <- read_tsv('data/deathtable.tsv')
names(deathdata) <- c('date', 'daily_deaths', 'cumulative_deaths')
deathdata <- deathdata %>% mutate(
  date = as.Date(paste0("2020/", date), format="%Y/%m/%d")
) %>%
  filter(date >= first_day)
ggplot(deathdata,
       aes(date, daily_deaths)) + geom_point() + geom_line() +
  geom_smooth() +
  labs(title="Deaths")

########## confirmed cases - last 7 days may be incomplete

casedata <- read_tsv('data/casetable.tsv')
names(casedata) <- c('date', 'daily_cases', 'cumulative_cases')
casedata <- casedata %>% mutate(
  date = as.Date(paste0("2020/", date), format="%Y/%m/%d")
) %>%
  filter(date >= first_day)
ggplot(casedata,
       aes(date, daily_cases)) + geom_point() + geom_line() +
  geom_smooth(data=head(casedata,-7)) +
  labs(title="Confirmed cases",
       subtitle="Last 7 days may have incomplete data")


########## tests - last 7 days may be incomplete

labdata <- read_tsv('data/labtable.tsv')
names(labdata) <- c('date', 'mn_lab_tests', 'ext_lab_tests',
                    'total_cumulative_tests')
labdata <- labdata %>% mutate(
  date = as.Date(paste0("2020/", date), format="%Y/%m/%d"),
  daily_tests = mn_lab_tests + ext_lab_tests
) %>%
  filter(date >= first_day)
ggplot(labdata, aes(date, daily_tests)) + geom_point() + geom_line() +
  geom_smooth(data=head(labdata,-7)) +
  labs(title="Tests",
       subtitle="Last 7 days may have incomplete data")
```

Data from https://www.health.state.mn.us/diseases/coronavirus/situation.html.

The graphs are over the last 8 weeks (56 days), to let us see more recent trends.

The trend lines end early for data that may be incomplete.

The smoothing is default geom_smooth, which in this case is loess (local estimation).
Roughly, it fits a low-degree polynomial to data points near the point in question.
hen the grey shaded area is a 95% confidence interval, which roughly means there's
a 95% chance the true mean is in that area.

It sounds fancy, but I just use the default settings for the tool (ggplot),
and it looks plausible.
