---
# title: "Minnesota Covid 19 data"
# date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
# https://stackoverflow.com/a/13516042
geometry: margin=1cm
---

# `r format(Sys.time(), '%B %e, %Y')`

```{r setup, include=FALSE}
library(tidyverse)
library(kableExtra)
library(scales)
# I thought lubridate came with tidyverse, but maybe not?
library(lubridate)

knitr::opts_chunk$set(
  # Don't put code in the doc
  echo = FALSE,
  # Don't put messages from code in the doc
  message = FALSE,
  # Don't put code warnings in the doc
  warning = FALSE,
  # stop on error instead of continuing
  error = FALSE,
  # Don't let figures float
  # Needs "float" package, see https://stackoverflow.com/a/36234023/34935
  fig.pos = "H",
  # Default height 3.75 inches: two per page plus text
  fig.height = 3.75
)

theme_set(theme_bw() +
            theme(plot.title=element_text(size=25),
                  panel.grid = element_line(colour="grey85")))
```

```{r funcs}
add_rolling_num <- function(df, col, window_days, suffix="_rolling") {
  c_col <- paste0('c_', col)
  df[[c_col]][!is.na(df[[col]])] <- cumsum(df[[col]][!is.na(df[[col]])])
  col1 <- paste0(col, suffix)
  df[[col1]] <- (
    df[[c_col]] - 
      # if lag of window_days is NA, pick the day before
      # let's hope there are not two NAs in a row
      coalesce(lag(df[[c_col]], window_days), lag(df[[c_col]], window_days+1))
  ) / window_days
  df
}

# add_rolling_rate <- function(df, rate, numer, denom, window_days) {
#   df <- add_rolling_num(df, numer, window_days)
#   df <- add_rolling_num(df, denom, window_days)
#   c_numer <- paste0('c_', numer)
#   c_denom <- paste0('c_', denom)
#   df[[rate]] <- (df[[c_numer]] - lag(df[[c_numer]], window_days)) /
#     (df[[c_denom]] - lag(df[[c_denom]], window_days))
#   df
# }

change_in_percent <- function(vec, window_days, round_n=0, unit='') {
  stopifnot(length(vec) >= window_days)
  last <- tail(vec, 1)
  last_minus <- coalesce(
    # data from window_days back:
    head(tail(vec, window_days+1), 1),
    # data from window_days-1 back (in case window_days back is NA)
    head(tail(vec, window_days+0), 1),
  )
  delta <- (last - last_minus)
  fmt <- paste0("%+.", round_n, "f", unit, " (%+.", round_n, "f%%)")
  list(delta=delta, percent=100*delta/last_minus,
       str=sprintf(fmt, delta, 100* delta / last_minus, 1))
}

CHANGE_WINDOW <- 14
ROLLING_WINDOW <- 7
```

```{r wastewater_graph}
wastewater_graph <- function() {
  wwdata <- read_csv('data/clean_load_data.csv')
  # wwdata <- add_rolling_num(wwdata, 'viral_load', ROLLING_WINDOW)
  viral_load_change <- change_in_percent(
    wwdata$copies_day_person_7day, CHANGE_WINDOW, unit='M')
  viral_load_percent <- viral_load_change$str

  p<-ggplot(wwdata %>% filter(date >= first_day),
         aes(date, copies_day_person_7day)) +
    geom_line(size=1) +
    geom_line(aes(y=copies_day_person_M_mn), alpha=0.25) +
    geom_point(aes(y=copies_day_person_M_mn), alpha=0.25) +
    labs(y="viral load, M copies/person/day",
         x=NULL,
         title='Wastewater viral load (metro)',
         subtitle=paste0('average ',
                  round(tail(wwdata$copies_day_person_7day, 1)),
                  "M, 14-day change ", viral_load_percent,
                  '. As of ', max(wwdata$date), '.',
                  '\nFrom the Metro treatment plant in Saint Paul,',
                  ' serving the 7-county metro.')
         ) +
    scale_x_date(date_minor_breaks="1 month", date_labels="%b %Y") +
    scale_size_identity() +
    scale_alpha_identity()

  return(list(data=wwdata, plot=p))
}
```

```{r vaccine_plot}
vaccine_plot <- function() {
  vax <- read_csv('data/vaxadminbyweek.csv') %>%
    rename(start_date=`Week Start Date`,
           num_administered=`Doses Administered Per Week`,
           total_administered=`Cumulative Total Doses Administered, by week`
           ) %>%
    mutate(
      start_date = parse_date(start_date, "%m/%d/%Y"),
      end_date = start_date + 7
    ) %>%
    filter(!is.na(start_date)) %>%
    filter(start_date >= first_day)
  vax_this_week <- vax %>% tail(1)
  p<-ggplot(vax, aes(start_date, num_administered)) +
    geom_bar(stat="identity") +
    scale_y_continuous(labels=comma) +
    scale_x_date(date_breaks="1 month", date_labels="%b") +
    labs(x="Week (so far)", y="# doses administered",
         title="Vaccines administered",
         subtitle=paste0(
           format(vax_this_week$num_administered, big.mark=",", scientific=F),
           " in the latest week, ",
           format(vax_this_week$total_administered, big.mark=",", scientific=F),
           " total so far.",
           " Includes current partial week."))
    # + theme(axis.text.x = element_text(angle = 20))

  return(list(plot=p, data=vax))
}
```


```{r per_county_graph}
rate_per_county_graph <- function(num_panels=4) {
  by_county_raw <- read_csv('data/county_rates.csv') %>%
    filter(state == "Minnesota")
  # last four dates
  last_five <- tail(sort(unique(by_county_raw$date_updated)), 5)
  by_county_raw <- by_county_raw %>% filter(date_updated %in% last_five)
  # add week number
  by_county <- by_county_raw %>%
    mutate(
      week_number = as.factor(week(date_updated)),
      county = gsub(" County", "", county),
      covid_cases_per_10k = covid_cases_per_100k / 10,
      covid_cases_per_10k_bucket = 
        cut(covid_cases_per_10k,
            c(-1,0,10,25,50,75,Inf), ordered_result = T)
    ) %>%
    select(county, covid_cases_per_10k, covid_cases_per_10k_bucket,
           date_updated, week_number)

  # Assign each county to a panel
  county_panel <- by_county %>% group_by(county) %>%
    summarize() %>%
    mutate(panel = floor((row_number()-1) / (length(county) / num_panels)))
  by_county <- by_county %>% merge(county_panel, by=c('county'))

  # rate level names:
  levels(by_county$covid_cases_per_10k_bucket) <-
    c("0", ">0 to <10", "10 to <25", "25 to <50",
      '50 to <75', '>=75')
  weeks <- by_county %>% group_by(week_number, date_updated) %>%
    summarize() %>%
    mutate(week_str = paste0(week_number, " ends ", date_updated))
  week_key_str <- paste0("Weeks: ", paste(weeks$week_str, collapse=", "))

  plot <- ggplot(
    by_county,
    aes(week_number, county,
        fill=covid_cases_per_10k_bucket)) +
    geom_tile() +
    scale_y_discrete(limits = rev) +
    scale_x_discrete(position="top") +
    labs(x=NULL, y=NULL,
         subtitle=paste0("Weekly case rate per 10,000 people, by county\n",
                         week_key_str)) +
    scale_fill_brewer(name="cases per 10k") +
    facet_wrap(~panel, scales="free_y", nrow=1) +
    theme(
      # remove facet_wrap headers
      # https://stackoverflow.com/a/10547999
      strip.background = element_blank(),
      strip.text.x = element_blank(),
      axis.text=element_text(size = rel(0.65)),
      legend.position="bottom",
      plot.subtitle=element_text(size = rel(0.8))
    )

  latest_week <- max(as.character(by_county$week_number))
  by_county_latest <- by_county %>% filter(week_number == latest_week)
  value1_summ <- by_county_latest %>%
    group_by(covid_cases_per_10k_bucket) %>% tally() %>% 
    arrange(desc(covid_cases_per_10k_bucket)) %>%
    mutate(perc = 100 * n / sum(n), cum_perc = 100 * cumsum(n) / sum(n))
  return(list(plot=plot,
              rate_hist=value1_summ,
              wcr_by_county_latest=by_county_latest,
              wcr_by_county=by_county))
}
```

```{r days_on_hand_graph}
days_on_hand_graph <- function() {
  daysonhand <- read_csv('data/DaysonHand_CCS_Chart_tcm1148-430853.csv') %>%
    rename(date=`Data Date`) %>%
    mutate(date = as.Date(date, format="%m/%d/%Y"))
  
  ggplot(daysonhand %>% filter(Location=="Hospital"),
         aes(date, `Days on Hand`,
             group=Item, color=Item)) + geom_point() + geom_line()
}
```

```{r graphs, fig.show = "hold", out.width = "50%"}
########## hospitalized

hospdata <- read_csv('data/h7day.csv') %>%
  select(-Rate, -total_pop) %>%
  pivot_wider(names_from=outcome, values_from=case_count, values_fill=0) %>%
  rename(all_admits=hospitalization, icu_admits=icu)
hospdata <- hospdata %>% mutate(
  date = as.Date(date, format="%m/%d/%Y"),
  gt_prev_date = date > lag(date)
)

# check we parsed the date correctly
stopifnot(hospdata$gt_prev_date %in% c(TRUE, NA))

last_day <- max(hospdata$date, na.rm=T)

# last 4 weeks
# first_day <- last_day - 28*1
# last 8 weeks
# first_day <- last_day - 28*2
# last 12 weeks
# first_day <- last_day - 28*3
# last 16 weeks
# first_day <- last_day - 28*4
# last 20 weeks
# first_day <- last_day - 28*5
# last 365 days
# first_day <- last_day - 365
# first_day <- as.Date('2021-03-01')
# first_day <- as.Date('2022-01-01')

first_day <- last_day - 364 - 28

# hospdata <- hospdata %>% filter(date <= last_day)

# Controls the amount of smoothing for the default loess smoother.
# Smaller numbers produce wigglier lines, larger numbers produce smoother lines.
# default: 0.75
loess_span <- 0.5

hospdata <- add_rolling_num(hospdata, 'icu_admits', ROLLING_WINDOW)
hospdata <- add_rolling_num(hospdata, 'all_admits', ROLLING_WINDOW)

hospdataf1 <- head(hospdata,-7)

# change in icu_admits_rolling over last CHANGE_WINDOW days
icu_admits_rolling_delta7_percent_l <- change_in_percent(hospdataf1$icu_admits_rolling, CHANGE_WINDOW)
icu_admits_rolling_delta7_percent <- icu_admits_rolling_delta7_percent_l$str

# change in all_admits_rolling over last CHANGE_WINDOW days
all_admits_rolling_delta7_percent_l <- change_in_percent(hospdataf1$all_admits_rolling, CHANGE_WINDOW)
all_admits_rolling_delta7_percent <- all_admits_rolling_delta7_percent_l$str

hospdata_long <- hospdata %>%
  pivot_longer(c('icu_admits', 'icu_admits_rolling', 'all_admits', 'all_admits_rolling')) %>%
  filter(date >= first_day) %>%
  mutate(color = ifelse(name %in% c('icu_admits', 'icu_admits_rolling'), 'icu_admits', 'all_admits'),
         size = ifelse(name %in% c('icu_admits', 'all_admits'), 0.25, 1)) %>%
  filter(!(name %in% c('icu_admits_rolling', 'all_admits_rolling') & (date > max(hospdataf1$date, na.rm=T))))

both_admits_plot <- function() {
  ggplot(hospdata_long %>% filter(name %in% c('icu_admits', 'all_admits')),
         aes(date, value, group=name, color=color)) +
    geom_point(alpha=0.25) +
    geom_line(data=hospdata_long %>% filter(name %in% c('icu_admits_rolling', 'all_admits_rolling')),
              aes(size=size, alpha=size)) +
    geom_line(data=hospdata_long %>% filter(name %in% c('icu_admits', 'all_admits'))
              %>% filter(date >= first_day),
              aes(size=size, alpha=size)) +
    labs(title="Hospital admissions",
         subtitle=paste0(
           "icu admits average ", round(tail(hospdata$icu_admits_rolling, 1)),
           ", 14-day change ", icu_admits_rolling_delta7_percent,
           "\nall admits average ",
           round(tail(hospdata$all_admits_rolling, 1)),
           ", 14-day change ", all_admits_rolling_delta7_percent, "\n",
           "Last 7 days may have incomplete data."),
         y="admitted on a given date.",
         x=NULL) +
    scale_y_continuous(limits=c(0,NA)) +
    scale_size_identity() +
    scale_alpha_identity() +
    scale_x_date(date_minor_breaks="1 month")
}
# both_admits_plot()

admits_plot <- function(title, admits_series_name, admits_rolling_series_name) {
  admits_series <- hospdata_long %>% filter(name %in% c(admits_series_name))
  admits_rolling_series <- hospdata_long %>% filter(name %in% c(admits_rolling_series_name))
  
  admits_rolling_delta7_percent_l <- change_in_percent(admits_rolling_series$value, CHANGE_WINDOW)

  ggplot(admits_series, aes(date, value)) +
    geom_point(alpha=0.25) +
    geom_line(data=admits_rolling_series, aes(size=size, alpha=size)) +
    geom_line(data=admits_series %>% filter(date >= first_day),
              aes(size=size, alpha=size)) +
    labs(title=title,
         subtitle=paste0(
           "average ", round(tail(admits_rolling_series$value, 1)),
           ", 14-day change ", admits_rolling_delta7_percent_l$str, ". ",
           "Last 7 days may have incomplete data."),
         y="admitted on a given date.") +
    scale_y_continuous(limits=c(0,NA)) +
    scale_size_identity() +
    scale_alpha_identity() +
    scale_x_date(date_minor_breaks="1 month")
}


# rm(hospdata)
# rm(hospdata_long)

########## deaths

deathdata <- read_csv('data/d7day.csv')

deathdata <- deathdata %>%
  mutate(
    # the_year = ifelse(row_number() <= 290, "2020", "2021"),
    # date = as.Date(paste0(the_year, "/", date), format="%Y/%m/%d")
    date = as.Date(date, format="%m/%d/%Y"),
    gt_prev_date = date > lag(date)
  ) %>%
  rename(daily_deaths = case_count) %>%
  add_rolling_num('daily_deaths', ROLLING_WINDOW)
# we got the year right:
stopifnot(deathdata$gt_prev_date %in% c(TRUE, NA))

# Make alpha of the last 7 days different
the_alpha <- 0.1
deathdata$alpha <- c(rep(0.25, nrow(deathdata)-7), rep(the_alpha, 7))

deathdataf <- deathdata # %>% filter(date >= first_day)
deathdataf1 <- head(deathdataf,-7)

change_in_daily_deaths_rolling_l <- change_in_percent(deathdataf1$daily_deaths_rolling, CHANGE_WINDOW)
latest_daily_deaths_rolling <- round(tail(deathdataf1$daily_deaths_rolling, 1))
latest_daily_deaths <- tail(deathdataf1$daily_deaths, 1)
death_plot <- function() {
  ggplot(deathdata %>% filter(date >= first_day),
         aes(date, daily_deaths)) +
    geom_point(aes(y=daily_deaths, alpha=alpha)) +
    geom_line(data=deathdataf1 %>% filter(date >= first_day),
              aes(y=daily_deaths_rolling), size=1) +
    geom_line(aes(y=daily_deaths, alpha=alpha)) +
    labs(title="Daily deaths",
         subtitle=paste0(
           "average ", latest_daily_deaths_rolling,
           ", 14-day change ",
           change_in_daily_deaths_rolling_l$str,
           ".  Last several days may have incomplete data."
           # "\n** As if March 9 had 2 deaths reported."
         ),
         x=NULL) +
    scale_y_continuous(limits=c(0,NA)) +
    scale_alpha_identity() +
    scale_x_date(date_minor_breaks="1 month", date_labels="%b %Y")
}
# death_plot()

########## confirmed cases - last 7 days may be incomplete

casedata <- read_csv('data/c7day.csv')
casedata <- casedata %>%
  mutate(
    date = as.Date(date, format="%m/%d/%Y"),
    gt_prev_date = date > lag(date),
    rn = row_number()
  ) %>%
  # ditch the date that can't be parsed
  filter(!is.na(date)) %>%
  add_rolling_num('case_count', ROLLING_WINDOW)
stopifnot(casedata$gt_prev_date %in% c(TRUE, NA))

# make alpha of last 7 days less
casedata$alpha <- c(rep(0.25, nrow(casedata)-7), rep(the_alpha, 7))
casedataf <- casedata # %>% filter(date >= first_day)
casedataf1 <- head(casedataf,-7)
change_in_cases_rolling_l <- change_in_percent(casedataf1$case_count_rolling, CHANGE_WINDOW)
latest_cases_rolling <- tail(casedataf1$case_count_rolling, 1)
case_plot <- function() {
  ggplot(casedataf %>% filter(date > first_day),
         aes(date, case_count)) +
    geom_point(aes(alpha=alpha)) +
    geom_line(aes(alpha=alpha)) +
    geom_line(data=casedataf1 %>% filter(date >= first_day),
              aes(y=case_count_rolling), size=1) +
    labs(title="Daily cases",
         subtitle=paste0(
           "average ", round(tail(casedataf1$case_count_rolling, 1)),
           ", 14-day change ",
           change_in_cases_rolling_l$str, ".",
           "  Last 7 days may have incomplete data."),
         x=NULL) +
    scale_y_continuous(limits=c(0,NA)) +
    scale_alpha_identity() +
    scale_x_date(date_minor_breaks="1 month", date_labels="%b %Y")
}

wastewater <- wastewater_graph()

case_plot()
death_plot()

admits_plot('ICU admissions', 'icu_admits', 'icu_admits_rolling')
admits_plot('All admissions', 'all_admits', 'all_admits_rolling')

vacc_result <- vaccine_plot()
wastewater$plot
vacc_result$plot
```

```{r case_rate_by_county, fig.height=3.15, fig.width=8}
rate_per_county_result <- rate_per_county_graph()
rate_per_county_result$plot
```

\newpage

## Latest data dates

```{r latest_dates}
dates <- data.frame(stringsAsFactors = F)
dates <- rbind(dates,
               hospdata_long %>% group_by(name) %>%
                 summarize(max_date=max(date)))
dates <- rbind(dates,
               data.frame(name="deaths", max_date=max(deathdata$date, na.rm=T)))
dates <- rbind(dates,
               data.frame(name="cases", max_date=max(casedataf$date)))

kable(dates)
```

# Introduction

What is the trend in Covid-19 cases in Minnesota?

Data from https://www.health.state.mn.us/diseases/coronavirus/situation.html
and https://mn.gov/covid19/data/response-prep/response-capacity.jsp.

Capacity csv is at https://mn.gov/covid19/assets/HospitalCapacity_HistoricCSV_tcm1148-449110.csv.

## 14-DAY CHANGES

```{r read_latest_cases}
dct <- read_tsv('data/dailycasetotal.tsv', col_names=c('label', 'num'))
todays_cases <- dct %>% filter(label=="Newly reported cases") %>% pull(num)
```


Uses confirmed data (all delayed 7 days except for deaths).

- hospital admissions: icu admits **up**
 `r sprintf("%+.0f%%", icu_admits_rolling_delta7_percent_l$percent)`,
 all admits **up** `r sprintf("%+.0f%%", all_admits_rolling_delta7_percent_l$percent)`

- daily deaths **up**
 `r sprintf("%+.0f", change_in_daily_deaths_rolling_l$delta)`
  (`r sprintf("%+.0f%%", change_in_daily_deaths_rolling_l$percent)`)
  to an average of `r latest_daily_deaths_rolling` as of `r tail(deathdataf1, 1)$date`. (Today: `r latest_daily_deaths`.)

- daily cases **up** `r sprintf("%+.0f%%", change_in_cases_rolling_l$percent)`,
  to **`r round(latest_cases_rolling)` per day** (7 days ago). (Today's cases: **`r todays_cases`**.)

## Per county summary

```{r per_county_summ}
kable(rate_per_county_result$rate_hist)
```

## PEAK SUMMARY

How far are the 7-day rolling averages from the peak, in numbers and days?

```{r peak_vars}

peak_list <- function(rolling_vec, date_vec) {
  idx <- which.max(rolling_vec)
  max_rolling <- rolling_vec[idx]
  #max_rolling_date <- date_vec[idx]
  max_rolling_days <- as.numeric(Sys.Date() - date_vec[idx], unit="days")
  last_val <- tail(rolling_vec, 1)
  percent_peak_rolling <- round(100 * last_val / max_rolling, 0)
  return(list(
    max_rolling=max_rolling,
    max_rolling_days=max_rolling_days,
    percent_peak_rolling=percent_peak_rolling,
    last_val=last_val
  ))
}

icu_admits_peak <- peak_list(hospdataf1$icu_admits_rolling, hospdataf1$date)

all_admits_peak <- peak_list(hospdataf1$all_admits_rolling, hospdataf1$date)

daily_deaths_peak <- peak_list(deathdataf1$daily_deaths_rolling, deathdataf1$date)

case_count_peak <- peak_list(casedataf1$case_count_rolling, casedataf1$date)

# No scientific notation:
options(scipen=10)
```

- Daily deaths: currently at `r daily_deaths_peak$percent_peak_rolling`% of peak (`r round(daily_deaths_peak$max_rolling, 0)` people,  `r daily_deaths_peak$max_rolling_days` days ago)

- ICU daily admissions: currently at `r icu_admits_peak$percent_peak_rolling`% of peak (`r round(icu_admits_peak$max_rolling, 0)` people,  `r icu_admits_peak$max_rolling_days` days ago)

- All daily admissions: currently at `r all_admits_peak$percent_peak_rolling`% of peak (`r round(all_admits_peak$max_rolling, 0)` people,  `r all_admits_peak$max_rolling_days` days ago)

- Daily cases: currently at `r case_count_peak$percent_peak_rolling`% of peak (`r round(case_count_peak$max_rolling, 0)` cases,  `r case_count_peak$max_rolling_days` days ago)


## More plots

```{r days_on_hand_plot}
# doesn't seem useful:
# days_on_hand_graph()
```

```{r vaccine_doses_per_day}
# vacc_data_daily <- read_tsv('data/vaccinated.tsv') %>%
#   mutate(total_doses_daily = total_doses - lag(total_doses))
```
